---
title: "ADS503 — Diabetes Readmission: EDA & Preprocessing"
author: "Selin Karadag, Jordyn Gonzales, Kamran Shirazi"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-tools: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document covers exploratory data analysis (EDA) and preprocessing for the **Diabetes 130-US Hospitals (1999–2008)** dataset. The goal is to understand the data structure, handle missing values, and prepare a clean dataset for predictive modeling of 30-day hospital readmissions.

**Response variable:** `readmitted_binary`

- `1` = Readmitted within 30 days
- `0` = Not readmitted within 30 days (\>30 or NO)

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

library(tidyverse)
library(corrplot)
```

------------------------------------------------------------------------

## 1. Data Loading

```{r}
#| label: load-data

diabetes <- read.csv("diabetic_data.csv")
mapping  <- read.csv("IDS_mapping.csv")
```

```{r}
#| label: data-check

dim(diabetes)
dim(mapping)
names(diabetes)
```

The dataset contains **`r nrow(diabetes)` rows** and **`r ncol(diabetes)` columns**.

------------------------------------------------------------------------

## 2. Response Variable

### 2.1 Raw Distribution

```{r}
#| label: response-raw

table(diabetes$readmitted)
prop.table(table(diabetes$readmitted))
```

### 2.2 Create Binary Target

Patients readmitted in fewer than 30 days are coded `1`; all others are coded `0`.

```{r}
#| label: create-target

diabetes$readmitted_binary <- ifelse(
  diabetes$readmitted == "<30",
  1,
  0
)

table(diabetes$readmitted_binary)
prop.table(table(diabetes$readmitted_binary))
```

The dataset is **imbalanced** — only \~`r round(mean(diabetes$readmitted_binary) * 100, 1)`% of encounters result in a 30-day readmission.

------------------------------------------------------------------------

## 3. Missing Value Analysis

### 3.1 Identify Missing Values

Missing values in this dataset are encoded as `"?"` rather than `NA`.

```{r}
#| label: missing-raw

sort(colSums(diabetes == "?"), decreasing = TRUE)
```

### 3.2 Calculate Missing Percentages

```{r}
#| label: missing-pct

missing_pct <- round(
  colSums(diabetes == "?") / nrow(diabetes) * 100,
  2
)

sort(missing_pct, decreasing = TRUE)
```

------------------------------------------------------------------------

## 4. Preprocessing

### 4.1 Drop High-Missing Columns

`weight`, `medical_specialty`, and `payer_code` exceed acceptable missing thresholds and are removed.

```{r}
#| label: drop-columns

diabetes_clean <- diabetes %>%
  select(-weight,
         -medical_specialty,
         -payer_code)

dim(diabetes_clean)
```

### 4.2 Remaining Missing Values

```{r}
#| label: missing-after-drop

sort(colSums(diabetes_clean == "?"), decreasing = TRUE)
```

### 4.3 Impute Remaining `?` as "Unknown"

```{r}
#| label: impute-unknown

diabetes_clean$race[diabetes_clean$race == "?"]     <- "Unknown"
diabetes_clean$diag_1[diabetes_clean$diag_1 == "?"] <- "Unknown"
diabetes_clean$diag_2[diabetes_clean$diag_2 == "?"] <- "Unknown"
diabetes_clean$diag_3[diabetes_clean$diag_3 == "?"] <- "Unknown"

sort(colSums(diabetes_clean == "?"), decreasing = TRUE)
```

### 4.4 Remove ID Columns

`encounter_id` and `patient_nbr` are identifiers with no predictive value.

```{r}
#| label: drop-ids

diabetes_clean <- diabetes_clean %>%
  select(-encounter_id,
         -patient_nbr)

dim(diabetes_clean)
```

The cleaned dataset has **`r nrow(diabetes_clean)` rows** and **`r ncol(diabetes_clean)` columns**.

------------------------------------------------------------------------

## 5. Correlation Analysis

Correlation matrix across all numeric features.

```{r}
#| label: correlation
#| fig-width: 8
#| fig-height: 7

numeric_vars <- diabetes_clean %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars)

corrplot(cor_matrix,
         method = "color",
         type   = "upper",
         tl.cex = 0.8)
```

------------------------------------------------------------------------

## 6. Predictor Exploration

Mean values of key numeric predictors split by readmission status, to identify the strongest signals before modeling.

```{r}
#| label: predictor-means

aggregate(number_inpatient ~ readmitted_binary,
          data = diabetes_clean, FUN = mean)

aggregate(number_emergency ~ readmitted_binary,
          data = diabetes_clean, FUN = mean)

aggregate(num_medications ~ readmitted_binary,
          data = diabetes_clean, FUN = mean)

aggregate(time_in_hospital ~ readmitted_binary,
          data = diabetes_clean, FUN = mean)

aggregate(number_diagnoses ~ readmitted_binary,
          data = diabetes_clean, FUN = mean)
```

------------------------------------------------------------------------

## 7. Exploratory Visualizations

### 7.1 Target Variable Distribution

```{r}
#| label: fig-target
#| fig-cap: "Distribution of 30-day readmission outcomes."

ggplot(diabetes_clean,
       aes(x = factor(readmitted_binary))) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of 30-Day Readmissions",
    x     = "Readmitted Within 30 Days (0 = No, 1 = Yes)",
    y     = "Count"
  )
```

### 7.2 Time in Hospital

```{r}
#| label: fig-time
#| fig-cap: "Distribution of length of stay."

ggplot(diabetes_clean,
       aes(x = time_in_hospital)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Time in Hospital",
    x     = "Days in Hospital",
    y     = "Count"
  )
```

### 7.3 Readmission Rate by Age Group

```{r}
#| label: fig-age
#| fig-cap: "Proportion of readmissions by age group."

ggplot(diabetes_clean,
       aes(x    = age,
           fill = factor(readmitted_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Readmission Rate by Age Group",
    x     = "Age Group",
    y     = "Proportion",
    fill  = "Readmitted"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 7.4 Prior Inpatient Visits vs. Readmission

```{r}
#| label: fig-inpatient
#| fig-cap: "Prior inpatient visits by readmission status."

ggplot(diabetes_clean,
       aes(x = factor(readmitted_binary),
           y = number_inpatient)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(
    title = "Prior Inpatient Visits by Readmission Status",
    x     = "Readmitted Within 30 Days (0 = No, 1 = Yes)",
    y     = "Number of Prior Inpatient Visits"
  )
```

### 7.5 Prior Emergency Visits vs. Readmission

```{r}
#| label: fig-emergency
#| fig-cap: "Prior emergency visits by readmission status."

ggplot(diabetes_clean,
       aes(x = factor(readmitted_binary),
           y = number_emergency)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(
    title = "Prior Emergency Visits by Readmission Status",
    x     = "Readmitted Within 30 Days (0 = No, 1 = Yes)",
    y     = "Number of Prior Emergency Visits"
  )
```

### 7.6 Number of Diagnoses vs. Readmission

```{r}
#| label: fig-diagnoses
#| fig-cap: "Number of diagnoses by readmission status."

ggplot(diabetes_clean,
       aes(x = factor(readmitted_binary),
           y = number_diagnoses)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(
    title = "Number of Diagnoses by Readmission Status",
    x     = "Readmitted Within 30 Days (0 = No, 1 = Yes)",
    y     = "Number of Diagnoses"
  )
```

### 7.7 Number of Medications vs. Readmission

```{r}
#| label: fig-medications
#| fig-cap: "Number of medications by readmission status."

ggplot(diabetes_clean,
       aes(x = factor(readmitted_binary),
           y = num_medications)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(
    title = "Number of Medications by Readmission Status",
    x     = "Readmitted Within 30 Days (0 = No, 1 = Yes)",
    y     = "Number of Medications"
  )
```

------------------------------------------------------------------------

## 8. Summary

The preprocessing pipeline:

1.  Encoded the response variable as a binary outcome (`readmitted_binary`)
2.  Removed three high-missingness columns: `weight`, `medical_specialty`, `payer_code`
3.  Replaced remaining `"?"` values with `"Unknown"` in `race`, `diag_1`, `diag_2`, `diag_3`
4.  Dropped identifier columns `encounter_id` and `patient_nbr`

The resulting `diabetes_clean` dataset is ready for feature engineering and predictive modeling.
